---
title: "Sentiment and Lexical Diversity Analysis of Song Lyrics: 1958 - 2021"
author: "Louis Magowan"
date: "24/01/2022"
output: html_document
---

**Sources **

+ [Song lyrics API](https://api.lyrics.ovh/v1/)
+ [IBM's Tone Analyzer API](https://www.ibm.com/uk-en/cloud/watson-tone-analyzer): Used to do sentiment analysis on the song lyrics, using a free trial account with the $200 USD credit they give. [Full description of IBM Tone Analyzer tones](https://cloud.ibm.com/docs/tone-analyzer?topic=tone-analyzer-utgpe)
+ [Kaggle dataset containing all of Billboard's top 100 songs](https://www.kaggle.com/dhruvildave/billboard-the-hot-100-songs)


**Load packages**
```{r packages, warning=FALSE}
packages <- c("tidyverse", "httr", 
              "quanteda", "quanteda.textstats",
              "quanteda.textmodels",
              "plotly", "ggthemes")

suppressMessages(invisible(lapply(packages, library, character.only=TRUE)))

# Suppress summarise info
options(dplyr.summarise.inform = FALSE)
```


# Load Data

## Get Billboard Top 100 Songs 1958 - 2021
```{r load_billboard}
# Get list of top 100 US songs since 1958 and process for lyrics API
# Read in Billboard hot 100 songs
billboard <- read.csv("billboard_hot_100.csv")
# Select needed cols
billboard <- billboard %>% select(date, artist, song)
# Drop all the repeated data
billboard <- billboard %>% distinct(artist, song, .keep_all = TRUE)
# Process the columns into URL encoding format
billboard$artist_url <- billboard$artist %>%
  str_replace_all(" ", "%20") %>%
  paste0("/")
billboard$song_url <- billboard$song %>%
  str_replace_all(" ", "%20") %>%
  paste0("/")
# Merge the two url encoded cols
billboard$url <- paste0(billboard$artist_url, billboard$song_url)

# Create column to add lyrics to
billboard$lyrics <- 0
```

## Get Lyrics

__Pull lyrics for each song in the billboard top 100s over the years with lyrics.ovh API. This took ~ 8 hours, so the results have been saved into a csv for knitting purposes. Many of the songs didn't return lyrics (45%), most likely due to inconsistencies in song and artist names- however the sample is still sufficiently large (16K songs, 1958 - 2021).__
```{r get_lyrics, eval=FALSE}
# Make requests to lyrics API
base_url <- "https://api.lyrics.ovh/v1/"

# Add counter to check calls are working with
counter = 0
for (i in 1:length(billboard$url)){

  # Get the url string to add to the base
  song_url <- billboard$url[i]
  # Construct the full request URL for that song
  full_url <- paste0(base_url, song_url)
  # Make request
  r <- GET(full_url)
  # Increment counter
  counter <- counter + 1
  # If the request was okay, add lyrics
  if (r$status_code == 200){
    # Need to process the data slightly, often they include this source line
    lyrics <- content(r)$lyrics %>%
      str_replace("Paroles de la chanson.*\r\n", "")
    # Add song's lyrics
    billboard$lyrics[i] <- lyrics
    # If rate-limited, break and display where limit happened
    # No rate-limit occurred in full 8 hours
  } else if (r$status_code == 429){
    cat("Rate-limited on: ", counter)
    break
  }
  # Add intermediate saves to store data in case something goes wrong
  # Also display the counter every 500 calls to check it's running
  if (counter %% 500 == 0){
    print(counter)
    write.csv(billboard, "lyrics_raw.csv", row.names = FALSE)
  }
}
# Save to a csv
write.csv(billboard, "lyrics_raw.csv", row.names = FALSE)
```

```{r process_lyrics}
# Read in the lyrics data
billboard <- read.csv("lyrics_raw.csv")
# Select needed columns
billboard <- billboard %>% select(date, artist, song, lyrics) %>% 
  # Remove songs that didn't return lyrics
  filter(lyrics != 0)
# Add year col for joining
billboard$year <- billboard$date %>% substr(1, 4)

# Rename date col
billboard <- billboard %>% rename(song_date = date)
```

## Sentiment Analysis with IBM Watson

__Use IBM Watson Tone Analyser to do sentiment analysis on each song: Many songs were unable to have their sentiment analysed, or else didn't produce strong results for any one sentiment. This also took a long time, and used up a lot of the $200 free credit they gave.__

```{r sentiment_api, eval=FALSE}
# Enter params for the request
ibm_key <- "mind_your_own_beeswax"
ibm_url <- "https://api.eu-gb.tone-analyzer.watson.cloud.ibm.com/instances/mind_your_own_beeswax_again/v3/tone"
ibm_version = "2017-09-21"

# Construct df to append IBM data to
tone_df <- data.frame()
# Add counter to check calls are working with
counter = 0
# Iterate through each song's lyrics, analysing sentiment
for (i in 1:length(billboard$lyrics)){
  # Get lyrics of current song
  current_song <-  billboard$lyrics[i]
  # Add lyrics to params for URL / request
  # The URL encoding of the lyrics is taken care of by the GET function
  params = list(
  version = '2017-09-21',
  text = current_song)
  # Increment counter
  counter <- counter + 1

  # Make the request
  r <- GET(url = ibm_url,
           query = params,
           authenticate('apikey', ibm_key))

   # If the request was okay, add lyrics
  if (r$status_code == 200){
    # Parse the data
    parsed_r <- fromJSON(rawToChar(r$content))
    # Merge tone data with billboard data into new df
    tone_df <- parsed_r$document_tone$tones %>%
      bind_cols(billboard[i, ]) %>%
      bind_rows(tone_df)

    # Check for auth errors etc
  } else if (r$status_code == 401) {
    cat("API requests stopped on number : ", counter,
        "\n Status: ", r$status_code, "\n")
    write.csv(tone_df, "tone_raw.csv", row.names = FALSE)
    break
  }
    # Add intermediate saves to store data in case something goes wrong
  # Also display the counter every 500 calls to check it's running
  if (counter %% 500 == 0){
    cat(counter, " Status: ", r$status_code)
    write.csv(tone_df, "tone_raw.csv", row.names = FALSE)
  }
}
# Save the data in a csv
write.csv(tone_df, "tone_raw.csv", row.names = FALSE)
```

```{r}
# Read in the saved sentiment analysis data
tone_df <- read.csv("tone_raw.csv")
# Drop the rows for which sentiment couldn't be analysed
tone_df <- tone_df[complete.cases(tone_df), ]
# Add year col for joining
tone_df$year <- tone_df$song_date %>% substr(1, 4)
# Select the needed columns and reorder
tone_df <-  tone_df %>% select(year, artist, song, tone_name, score)
```

## Merge Data and Process

```{r}
# Merge sentiment data with song data
all_df <- merge(tone_df, billboard, by = c("artist", "song", "year"))
# Make dates into date type
all_df$year <- all_df$year %>% paste0("-01-01") %>% as.Date()
all_df$song_date <- all_df$song_date %>% as.Date()
```


# Analyze Data

## Visualise Changes in Sentiment
__Plot trends in key sentiments over time. Lyrics appear to have become less joyful, angrier and slightly more fearful.__
```{r}
# Find the average sentiment scores by year
sent_df <- all_df %>% 
  # Just look at these 3 significant emotions
  filter(tone_name %in% c("Anger", "Joy", "Fear")) %>% 
  group_by(year, tone_name) %>% 
  summarise(mean_score = mean(score,
                              na.rm = TRUE))
# Plot the average sentiment scores over time
sentiment_plot <- ggplot(data = sent_df,
                         aes(x = year,
                             y = mean_score,
                             color = tone_name))  +
  # Add points
  geom_point(size = 0.75, alpha = 0.75) + 
  # Add linear trend lines
  geom_smooth(formula = "y ~ x", method = "lm") +
  
  # Add titles and labels
  labs(title = "Sentiments in Song Lyrics Over Time",
       color = "Sentiment") +
  ylab("Sentiment Strength") + 
  # Adjust theme
  theme(legend.title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text = element_text(face = "bold",
                                   size = 8),
        axis.ticks.y = element_blank(),
        plot.title = element_text(face = "bold"),
        text = element_text(family = "Sans")
        )

# Make into interactive graph
ggplotly(sentiment_plot)
```

[Group by decade](https://stackoverflow.com/questions/47799182/grouping-data-in-r-and-summing-by-decade)

__Plot key sentiments by decade.__
```{r}
# Facet bars
decade_sent <- all_df %>% 
  filter(tone_name %in% c("Joy", "Sadness", "Anger", "Fear")) %>%
  mutate(decade = floor(as.numeric(substr(year, 1, 4))/10)*10) %>% 
      group_by(decade, tone_name) %>% 
      summarise(mean_score = mean(score,
                                  na.rm = TRUE)) %>%
  # Remove incomplete decades
  filter(!decade %in% c(1950, 2020))
# Plot mean sentiment by decade
ggplot(data = decade_sent,
       aes(x = tone_name,
           y = mean_score,
           fill = tone_name)) +
  # Add bars
  geom_bar(stat = "identity") +
  # Annotate bars
  geom_text(aes(label = round(mean_score, 2)),
           position = position_dodge(width = 0.9),
           vjust=-0.5) + 
  # Facet by decade
  facet_wrap(~decade) +
  # Adjust graph limits
  coord_cartesian(ylim = c(0.55, 0.725)) + 

  # Add titles and labels
  labs(title = "Sentiments in Song Lyrics By Decade",
       fill = "Sentiment") +
  ylab("Sentiment Strength") + 
  # Adjust theme
  theme(legend.title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text = element_text(face = "bold",
                                   size = 8),
        axis.ticks.y = element_blank(),
        plot.title = element_text(face = "bold"),
        )
```
## Analyze Lexical Diversity

__Evaluate lexical diversity of lyrics over time.__
```{r}
# Make lyrics into a corpus
lyric_corpus <- corpus(billboard, text_field = "lyrics")

# Get number of tokens in each song
docvars(lyric_corpus)$ntoken <- lyric_corpus %>% ntoken()
# Filter out the songs with barely any lyrics
lyric_corpus <- corpus_subset(lyric_corpus, ntoken > 50)

# Tokenize lyrics and calculate lexical diveristy
lexdiv <- lyric_corpus %>%
  # Remove numbers and punctuation
  tokens(remove_punct = TRUE,
         remove_numbers = TRUE,
         remove_symbols = TRUE) %>%
  # Make all lowercase
  tokens_tolower() %>% 
  textstat_lexdiv(measure = c("TTR", "Maas"))

# Merge lexical diversity measures into docvars
lexdiv <- bind_cols(docvars(lyric_corpus), lexdiv[, 2:3])
# Make year into date
lexdiv$year <- lexdiv$year %>% paste0("-01-01") %>% as.Date()

# Make into long format for plotting
lexdiv <- lexdiv %>% pivot_longer(c(TTR, Maas),
                                names_to = "lexdiv_measure",
                                values_to = "lexdiv_value"
                                )

# Group by year
lexdiv_grouped <- lexdiv %>% group_by(year, lexdiv_measure) %>%
  summarise(lexdiv_value = mean(lexdiv_value, 
                       na.rm = TRUE))

```

## Visualise Changes in Lexical Diversity

__Both measures show a decline in lexical diversity over time. The steeper decline when using TTR may be a statistical artefact however (see below).__
```{r}
# Plot lexical diversity over time
ggplot(lexdiv_grouped, aes(x = year,
                           y = lexdiv_value,
                           color = lexdiv_measure)) + 
  # Add line
  geom_line(size = 1) +
  # Add trend lines
  geom_line(stat = "smooth",
            formula = "y~x",
            method = "lm",
            alpha = 0.5) +
  # Change ylabel
  scale_y_continuous(name = "TTR or Maas Index\n") +
   # Add titles and labels
  labs(title = "Average Lexical Diversity in Song Lyrics Over Time",
       color = "Lexical Diversity Measure") +
  xlab("") + 
  # Adjust theme
  theme_economist()
```

__TTR had decreased quite a lot. If you look at the average number of words in a song however, we see that this is likely to be partly down to an artefact of how TTR is calculated. TTR is unique tokens divided by total tokens- so longer pieces of text may tend towards less lexical diversity in terms of TTR. The Maas measure tries to compensate for varying text sample sizes and produces a more humble decline in lexical diversity over time. __

[Quanteda: Notes on lexical diversity (TTR, Maas and more)](https://quanteda.io/reference/textstat_lexdiv.html)
```{r}
# Group by year to get average number of tokens
ntoken <- lexdiv %>% group_by(year) %>% summarise(ntoken = mean(ntoken,
                                                                na.rm = TRUE))
# Plot average number of tokesn over time
ggplot(ntoken, aes(x = year,
                           y = ntoken,
                           )) + 
  # Add line
  geom_line(size = 0.8, color = "orange") +
  # Change ylab
  scale_y_continuous(name = "Number of Words in A Song\n") +
   # Add titles and labels
  labs(title = "Average Number of Words in a Song Over Time") +
  xlab("") + 
  # Add theme
  theme_economist()
```
